---
title: "Deaths from suicide in young people, aged 11-25 years"
output: pdf_document
date: "2024-11-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Title: The epidemic of suicide in Scotland, aged 11-25

##introduction 


For my project I will be trying to answer the question: 'Is death by suicide rising in young people in Scotland?'. Suicide in young people is a tragic consequence of poor mental health, poor mental health services, and social impact on young people. Analysising data from suicide rates could help us in the future to identify risk factors and descrease and prevent suicide. 


## Load packages 

We first need to start by loading packages needed for the 

```{r load-packages}
library(tidyverse)

```

## Data acquisition

The data I will be using in this report is the "Deaths by suicide in young people, aged 11-25 years" from Scotpho: https://scotland.shinyapps.io/ScotPHO_profiles_tool/

## Read in data 

Firstly the data is 'read in' as follows. The data I am using in this report contains a "period" column. This columns cells contain numbers that are a range of five years. To ease visual representation in my diagram I will be using the initial year the 'period' begins. Due to this code has been written to command only the 1-4 digits be used. 

```{r read-data}
Suicide_data <- read.csv("~/Documents/Trends_ScotPHO_data_extract_2024-10-21.csv")

Suicide_data <- Suicide_data %>%
  mutate(year_start = substr(trend_axis, 1, 4))
```

## Inspect Data

We use the glimpse function to initially inspect the data we have imported. 

```{r inspect_data}
glimpse(Suicide_data)
```

##Data cleaning and preperation

## Select trend_axis and measure 

The data we are using in this report has 10 columns, we are most interested in two to make our graph from. Due to this we need to select and isolate the nesicary columns. 

```{r select_trend_axis_and_measure}
Deaths_from_suicide_in_young_people_aged_11_to_25_years <- Suicide_data %>%
  select(`trend_axis`,`measure`) %>%
  rename( `Frequency` = `trend_axis`,
          `Crude rate per 100000 population` = `measure`)
head(Deaths_from_suicide_in_young_people_aged_11_to_25_years)

```

## Remove missing values 

Missing values are then taken out from our data.In this data there have been no missing values or unique values. 

```{r missing-values}
Deaths_from_suicide_in_young_people_aged_11_to_25_years$`5-year period` %>% unique()
Deaths_from_suicide_in_young_people_aged_11_to_25_years$`Crude rate per 100,000 population` %>% table()
Deaths_from_suicide_in_young_people_aged_11_to_25_years %>% summary()
```
## Check that data is tidy


```{r tidy-data}
Suidicde_data <- Suicide_data %>% 
  mutate(year_start = substr(trend_axis, 1, 4))

```

## Plot graph 



```{r plot_graph_two}
Suicide_data %>%
  ggplot(aes(x = year_start, y = measure)) +
  geom_col(fill = "yellow", colour = "black") +
  labs(
    title = "Crude Suicide Rates by Year Start",
    x = "Year Start",
    y = "Crude Rate per 100,000 Population"
  ) +
  theme_minimal()

```



## Graph type 3 

```{r trend_line}
Suicide_data %>%
  ggplot(aes(x = as.numeric(year_start), y = measure)) +  
  geom_col(fill = "yellow", colour = "black") +  
  geom_smooth(method = "lm", se = TRUE, colour = "blue", size = 1) + 
  labs(
    title = "Crude Suicide Rates by Year Start with Trend Line",
    x = "Year Start",
    y = "Crude Rate per 100,000 Population"
  ) +
  theme_minimal()
```

```{r Moving_trend_line}
Suicide_data %>%
  ggplot(aes(x = as.numeric(year_start), y = measure)) +  
  geom_col(fill = "yellow", colour = "black") +  
  geom_smooth(method = "lm", formula = y ~ poly(x, 4), se = TRUE, colour = "blue", size = 1) +  
  labs(
    title = "Crude Suicide Rates by Year Start with Polynomial Trend Line",
    x = "Year Start",
    y = "Crude Rate per 100,000 Population"
  ) +
  theme_minimal()
```



















